---
title: "Admissions Data Analysis"
author: "Barboza-Salerno"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    number-sections: false
    theme: sandstone
---

# Introduction

This code analyzes **admissions data**, focusing on **predicting SAT Math scores** using SAT Verbal scores and high school class size. The analysis employs **linear regression models**, including [simple linear regression](02-simple-regression.qmd), [multiple linear regression](03-multiple-regression.qmd), correlation assessments, and visualizations to explore the relationships between these variables.

Here is the [admissions data](data/admissions.sav) and you can download the script to run in R [here](admissions_R_ex.R)

# Load Necessary Libraries
```{r}
library(foreign)    # For reading .sav files
library(ggplot2)    # For visualization
library(dplyr)      # For data manipulation
library(gtsummary)  # For creating regression tables
library(officer)    # For exporting to Word
library(flextable)  # For table formatting
```

# Load and Inspect Data
```{r}
admissions <- read.spss("data/admissions.sav", to.data.frame = TRUE)
head(admissions) # Display first few rows
```

# Data Preprocessing
```{r}
admissions_clean <- admissions %>% 
  dplyr::select(SAT_math, SAT_verbal, HS_class_size) %>% 
  na.omit()
```

# Simple Linear Regression: SAT Math ~ SAT Verbal
```{r}
simple_model <- lm(SAT_math ~ SAT_verbal, data = admissions_clean)
summary(simple_model) # Regression summary
```

### Interpretation:
- **SAT Verbal has a positive effect on SAT Math scores** (Estimate = 0.45572, p < 0.001).
- **Model explains ~20.2% of variance in SAT Math scores** (Adjusted R² = 0.2011), meaning SAT Verbal alone is a moderate predictor.
- **Residuals appear reasonably distributed**, suggesting no major violations of normality assumptions.

# Compute Fitted Values and Residuals
```{r}
admissions_clean$fitted_simple <- fitted(simple_model)
admissions_clean$residuals_simple <- residuals(simple_model)
```

# Compute Standardized Values
```{r}
admissions_clean$std_fitted_simple <- scale(admissions_clean$fitted_simple)
admissions_clean$std_residuals_simple <- scale(admissions_clean$residuals_simple)
```

# Visualization: Residuals vs. Fitted
```{r}
ggplot(admissions_clean, aes(x = std_fitted_simple, y = std_residuals_simple)) +
  geom_point(color = "blue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Standardized Residuals vs. Standardized Fitted Values (Simple Model)",
       x = "Standardized Fitted Values", y = "Standardized Residuals")
```

### Interpretation:
- **Residuals are fairly symmetrically distributed** around zero, suggesting the linear model is appropriate.
- **No major patterns** in residuals indicate homoscedasticity.

# Histogram of Residuals
```{r}
ggplot(admissions_clean, aes(x = residuals_simple)) +
  geom_histogram(bins = 30, fill = "lightblue", color = "black") +
  labs(title = "Histogram of Residuals (Simple Model)", x = "Residuals", y = "Frequency")
```

### Interpretation:
- **Residuals follow a normal-like distribution**, supporting the assumption of normality.

# Correlation Analysis
```{r}
cor(admissions_clean[, c("SAT_math", "SAT_verbal", "fitted_simple", "residuals_simple")], use = "complete.obs")
```

### Interpretation:
- **Strong positive correlation** between SAT Math and SAT Verbal (r ≈ 0.45).
- **Fitted values highly correlate with SAT Verbal**, confirming its predictive power.

# Multiple Linear Regression: SAT Math ~ SAT Verbal + HS Class Size
```{r}
multiple_model <- lm(SAT_math ~ SAT_verbal + HS_class_size, data = admissions_clean)
summary(multiple_model)
```

### Interpretation:
- **HS Class Size has a small but significant positive effect** (Estimate = 0.06030, p < 0.001).
- **Explained variance increases to 22.4%**, showing a slight improvement over the simple model.

# Visualization of SAT Verbal Effect
```{r}
ggplot(admissions_clean, aes(x = SAT_verbal, y = SAT_math)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", col = "blue") +
  labs(title = "Effect of SAT Verbal on SAT Math", x = "SAT Verbal", y = "SAT Math")
```

# Generate Regression Summary Table
```{r}
mlr_table <- multiple_model %>% 
  tbl_regression(label = list(SAT_verbal = "SAT Verbal Score", HS_class_size = "High School Class Size")) %>%
  modify_caption("Table 1: Multiple Linear Regression Results") %>%
  add_n()
```

# Save Table to Word
```{r}
mlr_flextable <- as_flex_table(mlr_table)
doc <- read_docx()
doc <- body_add_flextable(doc, value = mlr_flextable)
print(doc, target = "MLR_Results.docx")
```

# Conclusion

### Key Findings:
1. **SAT Verbal scores significantly predict SAT Math scores**, explaining ~20% of the variance.
2. **Adding HS Class Size improves the model slightly**, raising the explained variance to 22.4%.
3. **Both models suggest a linear relationship**, with residuals showing normality and no major violations.

### Implications:
- Schools aiming to improve SAT Math scores may focus on strengthening verbal skills.
- Class size appears to play a small but relevant role in performance.
- Further research could explore additional predictors (e.g., socioeconomic factors, study habits).
