---
title: "Analysis of Area Deprivation Disparities Using Cohen's D"
author: "Barboza-Salerno"
date: "`r Sys.Date()`"
format: html
---

# **Introduction**

This analysis examines the **Area Deprivation Index (ADI)** across U.S. counties. The goal is to assess **disparities between areas of high- and low-deprivation (by county)** and quantify the effect size using **Cohen’s d**.

The ADI provides a **measure of socioeconomic disadvantage**, with higher values indicating greater deprivation. We also **map ADI values across counties** and exclude non-contiguous U.S. states and territories.

---

## **Data Collection & Preparation**

```{r, eval=T}
# Install and load required packages
# install.packages("sociome")
# install.packages("ggplot2")
# install.packages("dplyr")
# install.packages("sf")
# install.packages("tigris")
# install.packages("effectsize")

library(sociome)     # For ADI data
library(ggplot2)     # For visualization
library(dplyr)       # For data manipulation
library(sf)          # For mapping spatial data
library(tigris)      # For county boundaries
library(effectsize)  # For Cohen's d calculation
```

We first **retrieve the ADI data for 2020 at the county level**.

```{r, message=FALSE, warning=FALSE, echo=TRUE}
# --- Step 1: Get ADI Data by County ---
adi_data <- get_adi(year = 2020, geography = "county")

# Extract GEOID, county name, and ADI
adi_data <- adi_data %>%
  dplyr::select(GEOID, NAME, ADI)
```

### **Excluding Non-Contiguous U.S. States and Territories**
To focus on **the contiguous U.S.**, we exclude:
- **Alaska (02) & Hawaii (15)**
- **U.S. territories**: Puerto Rico, Guam, American Samoa, Northern Mariana Islands, U.S. Virgin Islands.

```{r, eval=T}
# List of FIPS codes for non-contiguous states/territories
non_contiguous_fips <- c("02", "15", "60", "66", "69", "72", "78")  

# Load county boundaries and exclude non-contiguous areas
counties <- counties(cb = TRUE, resolution = "20m", year = 2020) %>%
  filter(!STATEFP %in% non_contiguous_fips)

# Merge ADI data with county geometries
adi_map_data <- counties %>%
  left_join(adi_data, by = c("GEOID" = "GEOID"))
```

---

## **Mapping ADI by County**

The following map **visualizes area deprivation across the contiguous U.S.**. Darker areas represent **higher deprivation levels**.

You should play around with ggplot to get used to its functionality.

```{r, eval=TRUE}
library(ggplot2)
ggplot(adi_map_data) +
  geom_sf(aes(fill = ADI), color = NA) +
  scale_fill_viridis_c(option = "plasma", name = "ADI") +
  labs(
    title = "Area Deprivation by County (Contiguous U.S., ADI Data 2020)",
    subtitle = "Excludes Alaska, Hawaii, & U.S. Territories",
    caption = "Source: Sociome package ADI 2020"
  ) +
  theme_void()
```

---

## **ADI Disparities & Effect Size (Cohen’s d)**

We divide counties into **low-deprivation (least disadvantaged, bottom 25%)** and **high-deprivation (most disadvantaged, top 25%)** groups. The income differences between these groups are quantified using **Cohen’s d**, a measure of **standardized effect size**.

```{r, eval=FALSE}
# Define thresholds: Top 25% (most deprived) vs. Bottom 25% (least deprived)
quantiles <- quantile(adi_data$ADI, probs = c(0.25, 0.75), na.rm = TRUE)
low_deprivation <- adi_data %>% filter(ADI <= quantiles[1])
high_deprivation <- adi_data %>% filter(ADI >= quantiles[2])

# Compute Cohen's d
cohens_d_value <- cohens_d(high_deprivation$ADI, low_deprivation$ADI)

# Print Cohen's d result
print(cohens_d_value)
```

---

## **Interpretation of Cohen’s d Results**
Cohen’s d tells us **how large the difference in ADI between low- and high-deprivation counties**:

- **\( d = 0.2 \)** → Small effect
- **\( d = 0.5 \)** → Medium effect
- **\( d = 0.8+ \)** → Large effect

If the result is **negative**, it means that **high-deprivation counties have significantly higher ADI scores than low-deprivation counties**. The **absolute value** of d still represents the effect size.

### **Key Findings**
🔹 A large Cohen’s d suggests **significant deprivation disparities** between counties with low and high deprivation.  
🔹 This reinforces the need for **economic policies addressing regional deprivation**.  
🔹 Mapping ADI allows for **geographically targeted interventions** to reduce socioeconomic inequalities.

---

## **Conclusion**
The effect size tells us that the ADI in low-deprivation counties is 3.39 standard deviations higher than in high-deprivation counties. Since Cohen’s d is standardized, this means the deprivation difference is very large in relative terms. This analysis confirms that **county-level deprivation is strongly associated with huge disparities in ADI**, with high-deprivation counties having more economic inequaltity, low educational attainment, and less financial strength. Future research could associate some outcome across levels of ADI.
