---
title: "Logistic Regression Example"
format:
  html:
    toc: true
    number-sections: false
    theme: sandstone
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# The Dataset
The data is a `mixed variable` dataset containing 14 variables of 297 patients for their heart disease diagnosis. The data comes in an R package called [`kmed`](https://search.r-project.org/CRAN/refmans/kmed/html/heart.html) which you can read about on your own. Patients were diagnosed with heart disease in four classes. We will used this dataset to illustrate how multiple risk factors are related to a heart disease diagnosis, such as age (years), sex (FALSE = female; TRUE = male), chest pain (1 = typical angina, 2 = atypical angina, 3 = non-anginal pain, and 4 = asymptomatic) and thalach (max heart rate achieved).

```{r, warning=FALSE, message=FALSE}
library(kmed)
dat <- heart
```

```{r, warning=FALSE, message=FALSE}
# select variables
library(dplyr)
dat <- dat %>%
  dplyr::select(
    age,
    sex,
    cp,
    thalach,
    class
  )
# print dataset's structure
str(dat)
```

```{r, warning=FALSE, message=FALSE}
# rename variables
dat <- dat %>%
  dplyr::rename(
    chest_pain = cp,
    max_heartrate = thalach,
    heart_disease = class
  )
```

```{r, warning=FALSE, message=FALSE}
# recode sex
dat$sex <- factor(dat$sex,
  levels = c(FALSE, TRUE),
  labels = c("female", "male")
)
```

```{r, warning=FALSE, message=FALSE}
# recode chest_pain
dat$chest_pain <- factor(dat$chest_pain,
  levels = 1:4,
  labels = c("typical angina", "atypical angina", "non-anginal pain", "asymptomatic")
)
```

```{r, warning=FALSE, message=FALSE}
# recode heart_disease into 2 classes
dat$heart_disease <- ifelse(dat$heart_disease == 0,
  0,
  1
)
```

```{r, warning=FALSE, message=FALSE}
# set labels for heart_disease
dat$heart_disease <- factor(dat$heart_disease,
  levels = c(0, 1),
  labels = c("no disease", "disease")
)
```

```{r, warning=FALSE, message=FALSE}
levels(dat$heart_disease)
```

```{r, warning=FALSE, message=FALSE}
# save model
m1 <- glm(heart_disease ~ age,
  data = dat,
  family = "binomial"
)
```

```{r, warning=FALSE, message=FALSE}
summary(m1)
```

```{r, warning=FALSE, message=FALSE}
# OR for age
exp(coef(m1)["age"])
```

```{r, warning=FALSE, message=FALSE}
# prob(heart disease) for age = 0
exp(coef(m1)[1]) / (1 + exp(coef(m1)[1]))
```

```{r, warning=FALSE, message=FALSE}
# 95% CI for the OR for age
exp(confint(m1,
  parm = "age"
))
```
```{r, warning=FALSE, message=FALSE}
# predict probability to develop heart disease
pred <- predict(m1,
  newdata = data.frame(age = c(30)),
  type = "response"
)
```

```{r, warning=FALSE, message=FALSE}
pred
```

```{r, warning=FALSE, message=FALSE}
# predict probability to develop heart disease
pred <- predict(m1,
  newdata = data.frame(age = c(30)),
  type = "response",
  se = TRUE
)
```


```{r, warning=FALSE, message=FALSE}
pred$fit
```
```{r}
# 95% confidence interval for the prediction
lower <- pred$fit - (qnorm(0.975) * pred$se.fit)
upper <- pred$fit + (qnorm(0.975) * pred$se.fit)
c(lower, upper)
```
```{r}
# 95% confidence interval for the prediction
lower <- pred$fit - (1.96 * pred$se.fit)
upper <- pred$fit + (1.96 * pred$se.fit)
c(lower, upper)
```
```{r}
library(sjPlot)
library(ggplot2)
```

```{r, warning=FALSE, message=FALSE}
# plot
plot_model(m1,
  type = "pred",
  terms = "age"
) +
  labs(y = "Prob(heart disease)") + theme_bw()
```

```{r, warning=FALSE, message=FALSE}
# levels for sex
levels(dat$sex)
```

```{r}
# save model
m2 <- glm(heart_disease ~ sex,
  data = dat,
  family = "binomial"
)
```

```{r, warning=FALSE, message=FALSE}
summary(m2)
```

```{r, warning=FALSE, message=FALSE}
exp(coef(m2)["sexmale"])
```

```{r, warning=FALSE, message=FALSE}
# prob(disease) for sex = female
exp(coef(m2)[1]) / (1 + exp(coef(m2)[1]))
```

```{r, warning=FALSE, message=FALSE}
chisq.test(table(dat$heart_disease, dat$sex))
```
```{r}
# predict probability to develop heart disease
pred <- predict(m2,
  newdata = data.frame(sex = c("male")),
  type = "response"
)
pred
```
```{r, warning=FALSE, message=FALSE}
# plot
plot_model(m2,
  type = "pred",
  terms = "sex"
) +
  labs(y = "Prob(heart disease)")
```

```{r, warning=FALSE, message=FALSE}
# create data frame of new patient
new_patient <- data.frame(
  age = 32,
  sex = "female"
)
```

```{r, warning=FALSE, message=FALSE}
m3 <- glm(heart_disease ~ sex + age,
  data = dat,
  family = "binomial"
)
# predict probability to develop heart disease
pred <- predict(m3,
  newdata = new_patient,
  type = "response"
)
# print prediction
pred
```
```{r}
# 1. age, sex and chest pain on prob of disease
plot_model(m3,
  type = "pred",
  terms = c("age",  "sex"),
  ci.lvl = NA # remove confidence bands
) +
  labs(y = "Prob(heart disease)")
```

```{r}
tab_model(m3, m2,
  show.ci = FALSE, # remove CI
  show.aic = TRUE, # display AIC
  p.style = "numeric_stars" # display p-values and stars
)
```

```{r, warning=FALSE, message=FALSE}
library(pROC)
```

```{r, warning=FALSE, message=FALSE}
# save roc object
res <- roc(heart_disease ~ fitted(m3),
  data = dat
)
# plot ROC curve
ggroc(res, legacy.axes = TRUE)
```

```{r, warning=FALSE, message=FALSE}
res$auc
```

```{r}
# plot ROC curve with AUC in title
ggroc(res, legacy.axes = TRUE) +
  labs(title = paste0("AUC = ", round(res$auc, 2)))
```
```{r, warning=FALSE, message=FALSE}
library(gtsummary)
```

```{r, warning=FALSE, message=FALSE}
# print table of results
tbl_regression(m3, exponentiate = TRUE)
```
```{r, warning=FALSE, message=FALSE}
library(finalfit)
# set variables
dependent <- "heart_disease"
independent <- c("age", "sex")
independent_final <- c("age", "sex", "chest_pain")

dat %>% or_plot(dependent, independent,
  table_text_size = 3.5 # reduce text size
)
```


