---
title: "NSCAW I In-Class Exercise"
author: "Barboza-Salerno"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    number-sections: false
    theme: sandstone
---

# **Introduction**

This exercise explores **child aggression and trauma** using the **NSCAW I dataset**. Because this is designed for you to do, I supressed all of the output. The chunks of code give you a hints regarding the code that will answer each question. Create an R script and answer each question below. Cut and paste the chunk of code into your script, and fill it in so that it runs.

### **About the Data**
The **National Survey of Child and Adolescent Well-Being (NSCAW I)** is a **longitudinal study** that examines children involved in the child welfare system. It collects **child well-being indicators, caregiver characteristics, and service utilization data**. The dataset includes information on **trauma exposure, behavioral outcomes, and demographics**.

For this assignment, we focus on three key variables:

| Variable  | Type      | Description  |
|-----------|----------|----------------------|
| `tra1`    | Numeric  | Trauma: PTS T Score - This variable represents the **Post-Traumatic Stress (PTS) T-score** for the child. It measures trauma-related symptoms based on standardized assessments, with higher values indicating **greater levels of trauma exposure or distress**. Typically, scores above a certain threshold (e.g., 65-70) suggest clinically significant trauma symptoms.
| `bcagg1`  | Numeric  | Aggressive Behavior Standardized Score - This variable is a **standardized measure of aggressive behavior** in children, derived from behavioral assessments. Higher values suggest **greater levels of externalizing behaviors such as verbal or physical aggression**. Scores in the clinically significant range (e.g., above 65) may indicate the need for behavioral intervention.
| `ageY`    | Numeric  | Child age in years |

---

# **Instructions**

## **Step 1: Load the Data**
Make sure all of these libraries are installed or you will not be able to run the code. Also, make sure that your data set is in a **subdirectory** called `data` in the same directory as the one in which your `R script` is saved.

```{r, eval=FALSE}
# Load necessary libraries
library(foreign)    # For reading SPSS files
library(dplyr)      # For data manipulation
library(ggplot2)    # For visualization
library(gtsummary)  # For creating regression tables
library(officer)    # For exporting tables to Word
library(flextable)  # For table formatting

# Load NSCAW I dataset
nscaw <- read.spss("data/nscaw.sav", to.data.frame = TRUE)

# Inspect the first few rows
head(nscaw)
```
**Question:** What type of variables are included in the dataset?

To get a sense of the variables in the dataset [click here](https://www.ndacan.acf.hhs.gov/presentations/Overview_of_NSCAW_I_and_II.pdf).

---

## **Step 2: Select Variables & Handle Missing Data**
```{r, eval=FALSE}
# Select relevant variables
nscaw_clean <- nscaw %>% 
  dplyr::select(FILL THE VARIABLES IN HERE) 

# Remove missing values
nscaw_clean <- na.omit(nscaw_clean)

# View summary statistics
summary(nscaw_clean)
```
**Hint:** Look at the summary statistics—do any variables have extreme values?

---

## **Step 3: Compute Correlations**
```{r, eval=FALSE}
# Compute correlation matrix
cor(nscaw_clean, use = "complete.obs")
```
**Question:** Which variables are most strongly correlated? Does the direction make sense?

---

Run the regression of child aggression on trauma score and age

## **Step 4: Run a Multiple Regression Model**
```{r, eval=FALSE}
# Run regression: Predict aggression using trauma score and age
reg_model <- lm(WRITE THE CODE TO RUN THE REGRESSION)

# Display model summary
summary(reg_model)
```
**Hint:** Examine the coefficients and p-values. Is trauma a significant predictor of aggression?

---

## **Step 5: Compute Fitted Values & Residuals**
```{r, eval=FALSE}
# Compute fitted values and residuals
nscaw_clean$fitted <- fitted(PROVIDE THE CODE HERE)
nscaw_clean$residuals <- residuals(reg_model)

# Compute standardized residuals
nscaw_clean$std_residuals <- PROVIDE THE CODE HERE(nscaw_clean$residuals)
```
**Question:** Why do we standardize residuals? What does this help us understand?

---

## **Step 6: Visualize Residuals**
```{r, eval=FALSE}
# Plot histogram of residuals
ggplot(nscaw_clean, aes(x = residuals)) +
  PROVIDE THE CODE HERE(bins = 30, fill = "lightblue", color = "black") +
  labs(title = "Histogram of Residuals", x = "Residuals", y = "Frequency")
```
**Hint:** Look for skewness—are residuals normally distributed?

---

```{r, eval=FALSE}
# Plot standardized residuals vs. predicted values
ggplot(nscaw_clean, aes(x = fitted, y = std_residuals)) +
  geom_point(PROVIDE THE CODE HERE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(PROVIDE THE CODE HERE)
```
**Hint:** If you see patterns in the residuals, what assumption may be violated?

---

## **Step 7: Create & Export Regression Table**
```{r, eval=FALSE}
# Generate regression summary table
reg_table <- reg_model %>% 
  tbl_regression(
    label = list(tra1 = "PTS Score", bcagg1 = "Aggressive Behavior", ageY = "Age")
  ) %>% 
  modify_caption("Table 1: Regression Results for Aggression")

# Convert to flextable and export to Word
reg_flextable <- as_flex_table(reg_table)
doc <- read_docx()
doc <- body_add_flextable(doc, value = reg_flextable)
print(doc, target = "Regression_Results.docx")
```
**Hint:** Where does the p-value appear in the regression table? How does it inform significance?

---

## **Deliverables**

Discuss the following:

1. **Correlation matrix** output
2. **Regression model summary**
3. **Histogram of residuals**
4. **Residuals vs. fitted plot**
5. **Regression results table (Word document)**

---

## **Discussion Questions**
1. Based on your results, does **trauma significantly predict aggression**?
2. How does **child age influence the relationship between trauma and aggression**?
3. Are the **residuals normally distributed**? If not, what would you do differently?
4. If the **p-value for trauma is significant**, what does that imply in real-world terms?
5. How could you **improve this model**?

**Final Thought:** This exercise helps you develop a deeper understanding of how trauma influences child behavior. Future analyses could incorporate more variables (e.g., social support, environment) for better predictions.
